Environmental and geographic variables are effective surrogates for genetic variation in conservation planning

### 2016-07-22 code/rmarkdown/article.Rmd
+ fixes to make compatiable with pnas + change to raptr

---
title: "Are environmental and geographic variables effective surrogates for genetic variation in conservation planning?"
author: |
  | Jeffrey O. Hanson$^1$, Jonathan R. Rhodes$^2$, Cynthia Riginos$^1$, Hugh P. Possingham$^1$, Richard A. Fuller$^1$
  | $^1$School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  | $^2$School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
  | Correspondance should be addressed to jeffrey.hanson@uqconnect.edu.au
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: ["conservation", "biodiversity", "AFLP", "reserve-selection"]
abstract: "Insert abstract here."
output:
  rmarkdown::pdf_document:
    toc: false
    toc_depth: 5
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: preamble-latex.tex
      after_body: postamble-latex.tex
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
---

@@

# set options
options(error=NULL)
```

<!--  title page -->
Short running title: Genetic surrogates in conservation \newline
Abstract word count: XXXXX \newline
Main text word count: XXXXX \newline
Number of references: XXXX \newline
Number of equations: XXXX \newline
Number of figures: XXXX \newline
Number of tables: XXXX \newline
Number of text boxes: XXXX \newline
\newpage

## Introduction
<!--  main text -->
Protected areas spearhead conservation efforts [@r440]. These places buffer species from anthropogenic impacts and provide places for species to persist. However, the resources available for conservation are limited. Thus protected areas need to be sited in places that achieve conservation goals for minimal cost. To ensure that the plans for protected areas (prioritisations) are cost-effective, they are often generated using decision support tools to identify near-optimal solutions. These prioritisations often contain plans for a network of multiple protected areas so they can preserve biological processes (eg. connectivity between populations). For instance, \texttt{Marxan} [@r22] uses amount-based targets to identify protected area networks that preserve a suitable amount of habitat for each species (eg. 100 km $^2$ of land occupied by the species needs to be secured in a reserve) and weights to penalise fragmented solutions (ie. the \textit{BLM} parameter). However, conservation planning exercises typically assume that all individuals in the same species are equivalent.

Intra-specific genetic variation is important for long-term species persistence. As a consequence, there has been increasing interest in designing prioritisations that preserve and cultivate this variation [@r67; @r3; @r61]. Although the strength of natural selection is a continuous force; broadly speaking, genetic variation can be classified as either adaptive or neutral [reviewed in @r480]. Adaptive genetic variation is associated with loci that (significantly) affect fitness. Typically, such "adaptive" loci are detected due to anomalous patterns among individuals [eg. @r456; @r487; @r480; @r490; but see @r489 for alternative methods]. By preserving existing patterns of adaptive variation, protected areas can ensure that populations with particularly beneficial adaptations are not lost [@r3]. In contrast, neutral genetic variation is associated with loci that do not (significantly) affect fitness. Neutral variation often reflects the evolutionary history of different populations. By preserving existing patterns of neutral variation, protected areas can safeguard against the adverse effects of low genetic diversity [eg. inbreeding depression; @r67]. Thus protected area networks should aim to preserve both adaptive and neutral patterns of genetic variation for species of conservation interest [@r67; @r3]. Despite this, genetic data are not often used to inform conservation planning exercises because such data require considerable investment to obtain and analyze [@r61; but see @r5 and @r157 for developments on using genetic data in conservation planning]--especially since most large-scale exercises involve hundreds of species. Thus prioritisations may fail to secure a sufficient proportion of the intra-specific genetic variation of  species, and ultimately fail to deliver long-term conservation outcomes.

Recently, conservation scientists have been investigating the use of surrogate data to generate prioritisations that adequately preserve intra-specific genetic variation--but without actually needing to utilise genetic data to acheive this. Since adaptive genetic variation is ultimately driven by extrinsic selection pressures, environmental variables have been used to guide reserve selection as surrogates for adaptive genetic variation [eg. environmental conditions; @r247; @r2]. By preserving a more representative sample of environmental conditions across the species' ranges, these researchers aimed to generate prioritisations that would preserve more adaptive variation. However, the effectiveness of this approach remains unverified. On the other hand, neutral genetic variation arises from a reduction in gene flow between individuals. Surrogates for neutral variation have been based on variables that predict the level of connectivity between planning units. For instance, the isolation by distance theory [IBD; @r259] predicts that populations further apart will have less connectivity, exchange less genetic material, and in turn have less genetic variation in common. Based on this theory, geographic distance-based surrogates have been found to improve prioritisations in an island network [@r200]. However this remains untested in spatially contiguous systems where connectivity is complicated by additional factors [eg. terrain or habitat type; @r474; @r497]--the very systems commonly featured in real-world conservation planning exercises [eg. @r197]. Before conservation planners can reliably use surrogates for genetic data in conservation planning, the effectiveness of these surrogates need to be verified. Otherwise, conservation planners may waste precious resources on protecting places that provide little benefit to biodiversity.

Here we investigate the effectiveness of environmental and geographic surrogates for adaptive and neutral genetic variation in the context of conservation planning. First, we aim to verify if environmental and geographic variables correlate with broad-scale patterns in adaptive and neutral variation (respectively). Although these variables may correlate with genetic variation, they may not be effective surrogates if they have poor predictive ability or if these correlations are driven by spatial auto-correlation. Second, we aim to verify that environmental and geographic variables are effective surrogates for adaptive and neutral variation (respectively). We expect that prioritisations generated using surrogate-based targets will secure more genetic variation than prioritisations generated using only amount-based targets with the same number of planning units. Third, we aim to determine if the addition of surrogate-based targets to conservation planning exercises can result in more effective prioritisations. We expect that prioritisations generated using amount- and surrogate-based targets will sufficiently preserve the genetic variation of more species than those generated using only amount-based targets. Furthermore, we aim to test this hypothesis under several of different scenarios to see whether the additional constraints of real-world planning exercises exacerbate or nullify this effect. If these surrogates can improve the intra-specific genetic variation captured in a prioritisation, then conservation planners may use these surrogates when genetic data is not available, and in turn deliver more effective protected areas.

## Methods
### Study system
We used species distribution and genomic data collected by the international IntraBioDiv project in the European Alps [@r451; see @r452 and @r478 for further explanation of data collection methods]. This data set contains information for 27 alpine plant species (Figure 1a). It was collected using a $20^{\prime}$ longitude by $21^{\prime}$ latitude grid (approx. 22.3 km $\times$ 25 km; Figure S1). Project members visited every second grid cell, and if a species was detected in a cell, samples were collected from three individuals. They genotyped samples using amplified fragment length polymorphisms [AFLP; @r453], and constructed matrices denoting the presence/absence of polymorphisms at loci for each species (mean `r round(mean(sapply(spp.StructureData.LST, structurer:::n.loci)),2)` $\pm$ `r round(sd(sapply(spp.StructureData.LST, structurer:::n.loci)),1)` SD markers genotyped per species; see Table 1 in @r478 for the number markers for each species). Thus the data set contains information describing the genomic properties of individuals for each species present in every second grid cell.

>>>>

---
title: "Environmental and geographic variables are effective surrogates for genetic variation in conservation planning"
author:
  - name: Jeffrey O. Hanson
    affiliation: a,1
  - name: Jonathan R. Rhodes
    affiliation: b
  - name: Cynthia Riginos
    affiliation: a
  - name: Hugh P. Possingham
    affiliation: a
  - name: Richard A. Fuller
    affiliation: a
address:
  - code: a
    address: School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  - code: b
    address: School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
corresponding_author:
  - code: 1
  - text: "To whom correspondence should be addresssed. E-mail: jeffrey.hanson@uqconnect.edu.au"
author_contributions: |
  All authors developed the study and drafted the manuscript. JOH performed all analyses.
lead_author_surname: "Hanson"
abstract: |
  Please provide an abstract of no more than 250 words in a single paragraph. Abstracts should explain to the general reader the major contributions of the article. References in the abstract must be cited in full within the abstract itself and cited in the text.
significance: |
  Protected areas are used to preserve biodiversity patterns and processes. Since the resources available for conservation are limited, protected areas need to be sited in cost-effective locations. However, despite the stated emphasis on biodiversity processes, evolutionary processes are not often used to guide reserve selection. This is due to the fact that genetic data is often too costly to obtain and process. Here we found that environmental and geographic variables are cost-effective surrogates for genetic data. These variable can be used to identify candidate reserve systems that secure the adaptive and neutral evolutionary processes for a greater number of species. This research highlights the importance of obtaining a representative sample of species when generating broad-scale prioritisations.
acknowledgements: |
  JOH is funded by an Australian Postgraduate Award (APA) scholarship. RAF has an Australian Research Council Future Fellowship. This work was supported by the Centre of Excellence for Environmental Decisions (CEED) and the Landscape Ecology and Conservation Group (LEC) at The University of Queensland.
keywords:
  - conservation
  - biodiversity
  - AFLP
  - genetic diversity
pnas_type: pnasresearcharticle
bibliography: references.bib
csl: reference-style.csl
output:
  rticles::pnas_article:
    includes:
      in_header: [preamble.tex]
---

@@

# set options
options(error=NULL)

# download basemap
data(countriesHigh)
countries.FPLY <- countriesHigh[
	countriesHigh$ADMIN %in% c(
		'Italy', 'Switzerland', 'France', 'Austria',
		'Germany', 'Slovenia', 'Croatia', 'Hungary',
		'Monaco', 'Germany', 'Slovakia', 'Czech Republic'
	)
,] %>% spFortify
grid.FPLY <- spFortify(grid.PLY)
grid.sub.FPLY <- spFortify(grid.sub.PLY)

# define ggplot2 color wheel function
gg_color_hue <- function(n) {
	hues <- seq(15, 375, length=n+1)
	hcl(h=hues, l=65, c=100)[1:n]
}
```

<!--  Introduction -->
Protected areas spearhead conservation efforts [@r440]. These places buffer species from anthropogenic impacts and provide places for species to persist. However, the resources available for conservation are limited. Thus protected areas need to be sited in places that achieve conservation goals for minimal cost. To ensure that the plans for protected areas (prioritisations) are cost-effective, they are often generated using decision support tools to identify near-optimal solutions. These prioritisations often contain plans for a network of multiple protected areas so they can preserve biological processes (eg. connectivity between populations). For example, \texttt{Marxan} [@r22] uses amount-based targets that specify a minimum amount of habitat to preserve for each species to identify cost-effective solutions (eg. 100 km $^2$ of land occupied by the species needs to be secured in a reserve). To promote connecitivty among multiple reserves, \texttt{Marxan} also contains the functionality to penalise fragmented solutions (using the boundary length modifier (\textit{BLM}) parameter). However, conservation planning exercises typically assume that all individuals in the same species are equivalent.

Intra-specific genetic variation is important for long-term species persistence. As a consequence, there has been increasing interest in designing prioritisations that preserve and cultivate this aspect of biodiversity [@r67; @r3; @r61]. Although the strength of natural selection is a continuous force; broadly speaking, genetic variation can be classified as either adaptive or neutral [reviewed in @r480]. Adaptive genetic variation is associated with loci that (significantly) affect fitness. Typically, such "adaptive" loci are detected due to anomalous patterns among individuals [eg. @r456; @r487; @r480; @r490], though phenotypic-genetic associations such as quantitative trait analyses or genome-wide association measures are more likely to identify functionally important genomic regions for individual species [@rXXXX; @rXXXX]. By preserving existing patterns of adaptive variation, protected areas can ensure that populations with particularly beneficial adaptations are not lost [@r3]. In contrast, neutral genetic variation is associated with loci that do not (significantly) affect fitness. Neutral variation often reflects the evolutionary history of different populations. By preserving existing patterns of neutral variation, protected areas can safeguard against the adverse effects of low genetic diversity [eg. inbreeding depression; @r67]. Thus protected area networks should aim to preserve both adaptive and neutral patterns of genetic variation for species of conservation interest [@r67; @r3]. Despite this, genetic data are not often used to inform conservation planning exercises because such data require considerable investment to obtain and analyze for a single species [@r61; but see @r5 and @r157 for developments on using genetic data in conservation planning], and most large-scale planning exercises involve hundreds of species [eg. @rXXX]. Thus prioritisations may fail to secure a sufficient proportion of the intra-specific genetic variation of species, and ultimately fail to deliver long-term conservation outcomes.

Recently, conservation scientists have proposed using surrogate data to generate prioritisations that adequately preserve intra-specific genetic variation--without needing to utilise genetic data to acheive this--but have yet to comprehensively assess whether the proposed spatial surrogates adequately represent relevant aspects of intra-specific genetic variation. Since adaptive genetic variation is ultimately driven by extrinsic selection pressures, environmental variables have been proposed to guide reserve selection as surrogates for adaptive genetic variation [eg. environmental conditions; @r247; @r2]. By preserving a more representative sample of environmental conditions across the species' ranges, these researchers aimed to generate prioritisations that would preserve more adaptive variation. However, the effectiveness of this approach remains unverified. On the other hand, neutral genetic variation arises from a reduction in gene flow between individuals. Over the last few years, the field of phylogeography has predominantly focused on describing neutral genetic diversity, and so the landscape factors that affect neutral variation are relatively well understood [@rXXXX]. Potential surrogates for neutral variation have been based on variables that predict the levels of connectivity between planning units. For instance, the isolation by distance theory [IBD; @r259] predicts that populations further apart will have less connectivity, exchange less genetic material, and in turn have less genetic variation in common. Based on this theory, geographic distance-based surrogates have been found to improve prioritisations in an island network [@r200]. However this remains untested in spatially contiguous systems where connectivity is complicated by additional factors [eg. terrain or habitat type; @r474; @r497]--the very systems commonly featured in real-world conservation planning exercises [eg. @r197]. Before conservation planners can reliably use surrogates for genetic data in conservation planning, the effectiveness of these surrogates need to be verified. Otherwise, conservation planners may waste precious resources on protecting places that provide little benefit to biodiversity.

Here we investigated the effectiveness of environmental and geographic surrogates for adaptive and neutral genetic variation in the context of conservation planning. To achieve this, we developed a novel framework for incorporating adaptive and neutral genetic data into conservation planning. First, we aim to verify if the spatial patterns in environmental and geographic variation correlate with the broad-scale spatial patterns in adaptive and neutral variation (respectively). Second, we aim to test whether environmental and geographic variables are effective surrogates for adaptive and neutral variation (respectively) when generating prioritisations. We expect that prioritisations generated using surrogate-based targets will secure more genetic variation than prioritisations generated using only amount-based targets with the same number of planning units. Third, we aim to determine if the addition of surrogate-based targets to conservation planning exercises can result in more effective prioritisations. We expect that prioritisations generated using amount- and surrogate-based targets will sufficiently preserve the genetic variation of more species than those generated using only amount-based targets. Furthermore, we aim to test this hypothesis under several of different scenarios to see whether the additional constraints of real-world planning exercises exacerbate or nullify this effect. If these surrogates can improve the intra-specific genetic variation captured in a prioritisation, then conservation planners may use these surrogates when genetic data is not available, and in turn deliver more effective protected areas.

Results {.unnumbered}
=====================

<!-- summary of methods -->
We successfully mapped the main spatial patterns of adaptive (if present) and neutral variation for each of the 27 species. The outlier detection methods detected putatively adaptive genetic variation in `r as.character(as.english(length(env.surrogacy.LST)))` species. Generally, the two outlier detection methods gave fairly consistent results (mean `r round(mean(spp.outlier.methods.congruence, na.rm=TRUE),2)` % loci per species $\pm$ `r round(sd(spp.outlier.methods.congruence, na.rm=TRUE),2)` SD). Of the species that were associated with adaptive genetic variation, only a few loci were detected as under selection (mean `r round(mean(spp.outlier.loci.props, na.rm=TRUE),2)` % loci per species $\pm$ `r round(sd(spp.outlier.loci.props, na.rm=TRUE),2)` SD). After identifying the loci showing strong signals of selection, we used NMDS analyses to identify the main gradients of adaptive (if detected) and neutral variation for each species. For all species in which we detected outlier loci, only a couple of continuous dimensions were needed to sufficiently describe their patterns of adaptive genetic variation (all $K = `r unique(na.omit(adaptive.k))`$; Table S2). On the other hand, a few additional continuous dimensions were occasionally needed to describe their patterns of neutral genetic variation (minimum $K = `r min(neutral.k)`$; mean $K = `r round(mean(neutral.k),2)`$; maximum $K = `r max(neutral.k)`$; Table S2). The ordinations resulting from the NMDS analyses were used to construct an adaptive (if present) and neutral genetic space for each species. The spatial distribution of these genetic spaces (Figures S29--S54) generally showed patterns of spatial auto-correlation (eg. _Cerastium uniflorum_, Figure SXX; _Dryas octopetala_, Figures SXXX)--a good sign that the environmental and geographic variables may prove effective surrogates.

<!-- correlation analysis -->
The correlation between dissimilarity matrices based on the environmental and geographic characteristics of the planning units and dissimilarity matrices based on the adaptive and neutral genetic characteristics of individuals in the planning units (Table 1). Specifically, environmental variation was found to significantly correlate with adaptive genetic variation for `r as.character(as.english(sum(correlation.model.summary.DF$Environmental.corr.P<0.05,na.rm=TRUE)))` species of the `r as.character(as.english(sum(!is.na(correlation.model.summary.DF$Environmental.corr.P))))` species associated with adaptive variation (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` SD for significant models). These results suggest that prioritisations with different environmental characteristics tend to have individuals with different adaptive genetic characteristics. Additionally, geographic distances between planning units were found to significantly correlate with neutral genetic variation among planning units for `r sum(correlation.model.summary.DF$Geographic.corr.P<0.05,na.rm=TRUE)` species of the total `r sum(!is.na(correlation.model.summary.DF$Geographic.corr.P))` (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` SD for significant models). Thus planning units that were located further apart tended to have individuals with increasingly different neutral genetic characteristics. After verifying that the geographic and environmental variables generally correlated with genetic variation, our next step was to determine how effective these variables were as surrogates in the context of conservation planning.
```{r, echo=FALSE, results='asis'}
curr.file <- tempfile(fileext='.tex')
x <- correlation.model.summary.DF %>%
	select(-contains('raw')) %>%
	mutate(
		Environmental.corr.P=pvalString(Environmental.corr.P),
		Geographic.corr.P=pvalString(Geographic.corr.P),
		Geographic.Marginal.R2=pvalString(Geographic.Marginal.R2),
		Geographic.Conditional.R2=pvalString(Geographic.Conditional.R2),
		Environmental.Marginal.R2=pvalString(Environmental.Marginal.R2),
		Environmental.Conditional.R2=pvalString(Environmental.Conditional.R2),
		blank=rep('', length(Species)),
		Species=paste0('\\textit{', gsub('\\ ', ' ', Species), '}')) %>%
	mutate(Environmental.Chisq=sapply(round(Environmental.Chisq,2), as.character)) %>%
	mutate(Geographic.Chisq=sapply(round(Geographic.Chisq,2), as.character)) %>%
	select(Species:Environmental.corr.P, Geographic.Marginal.R2:Geographic.corr.P) %>%
	latex(file=curr.file, 
		cgroup=c('', 'Environmental', 'Geographic'), n.cgroup=c(1, 4, 4), col.just=rep('c', 9),
		digits=2, rowname=NULL, first.hline.double=FALSE, booktabs=TRUE, here=TRUE,
		colheads=c('\\bfseries{Species}', '$MR^2$', '$CR^2$', '$\\chi^2$', '\\textit{P}', '$MR^2$', '$CR^2$', '$\\chi^2$', '\\textit{P}'),
		caption='Summary of maximum likelihood population-effects (MLPE) models. Briefly, for each species associated with potentially adaptive loci, a model was fit to correlate dissimilarities among planning units where the species in terms of the adaptive genetic characteristics of individuals in them with dissimilarities in terms of the environmental conditions inside the planning units (denoted in the environmental column group). Similarly, for each species, a model was fit to correlate dissimilarities among the planning units in terms of the neutral genetic characteristics of indiviudals in them with dissimilarities in terms of the geographic position of the planning units (denoted in the geographic column group). Data shows the results from comparing each model to its corresponding null model using $\\mathbf{\\chi^2}$ tests (1 degree of freedom), and the $\\mathbf{R^2}$ ($\\mathbf{MR^2}$) and conditional $\\mathbf{R^2}$ ($\\mathbf{CR^2}$) for each model.'
	)
readLines(curr.file)[-1] %>% paste0(collapse='\n') %>% 
	gsub(pattern='\\cline{1-', replacement='\\cline{2-6} \\cline{8-', fixed=TRUE) %>% 
	gsub(pattern='{c}{\\textit{P}}\\tabularnewline', replacement='{c}{\\textit{P}}\\T\\B\\tabularnewline', fixed=TRUE) %>% 
	gsub(pattern='\\toprule', replacement='', fixed=TRUE) %>%
	gsub(pattern='\\midrule', replacement='\\midrule', fixed=TRUE) %>%
	gsub(pattern='\\bottomrule', replacement='\\midrule', fixed=TRUE) %>%
	cat()
```

<!-- surrogacy analysis -->
The environmental and geographic variables were moderately effective surrogates for genetic variation in most species (Figure 1). The relationship between the proportion of adaptive genetic variation and environmental variation secured in a given prioritisation was found to differ depending on the species for which the prioritisation was generated (species $\times$ proportion of environmental variation interaction term; ${\chi^2}_{`r env.surrogacy.AOV$df[[4]]`} = {`r round(env.surrogacy.AOV$Resid..Dev[[4]],2)`}; P `r pvalString(env.surrogacy.AOV$Pr..Chi.[[4]])`$). The slopes for each species were all significantly different to zero (minimum \textit{z} = `r round(max(env.surrogate.spp.DF$statistic),2)`; maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; Table S3). Similarly, the relationship between the proportion of neutral genetic variation and geographic variation secured in a prioritisation was also found to affected by the interaction between species and the proportion of neutral variation secured in the prioritisation (${\chi^2}_{`r geo.surrogacy.AOV$df[[4]]`} = `r round(geo.surrogacy.AOV$Resid..Dev[[4]],2)`$; $P$ `r pvalString(geo.surrogacy.AOV$Pr..Chi.[[4]])`). The slope for each species were also significantly different to zero (minimum \textit{z} = `r round(min(geo.surrogate.spp.DF$statistic),2)`; maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; Table S3). After establishing a relationship between the proportion of the surrogate variables and proportion of genetic variation secured in a prioritisation, the next step was to determine if prioritisations benefited from the addition of surrogate-based targets and verify if such benefits were evident across a variety of planning scenarios.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4.5, out.width='3.42in', out.height='2.5in', fig.cap='The relationship between surrogates and genetic variation secured in prioritizations. Panel (a) shows the relationship between the proportion of intra-specific adaptive genetic variation and environmental variation secured in a prioritization. Similarly, panel (b) shows the relationship between the proportion of intra-specific neutral variaiton and geograpgic variation secured in a prioritization. Panels (c) and (d) show the distribution of $R^2$ values for these models.'}
# make figure
p1 <- ggplot(aes(x=surrogate.held,y=genetic.held), data=env.surrogacy.DF) +
	xlab('Environmental variation\nsecured (%)') + 
	ylab('Adaptive genetic\nvariation secured (%)') +
	theme_classic() + theme(axis.line.x=element_line(color='black'),
		axis.line.y=element_line(color='black')) +
	geom_text(x=0.05, y=0.95, label='(a)') +
	coord_cartesian(xlim=c(0, 1))
for (i in seq_along(env.surrogacy.LST))
	p1 <- p1 + stat_smooth(method='glm', method.args=list(family='binomial'),
		data=filter(env.surrogacy.DF, Species==unique(Species)[i]),
		fullrange=TRUE, se=FALSE, colour='black')

p2 <- ggplot(aes(x=surrogate.held,y=genetic.held), data=geo.surrogacy.DF) +
	xlab('Geographic variation\nsecured (%)') + 
	ylab('Neutral genetic\nvariation secured (%)') +
	theme_classic() + theme(axis.line.x=element_line(color='black'),
		axis.line.y=element_line(color='black', colour='transparent')) +
	geom_text(x=0.05, y=0.95, label='(b)') +
	coord_cartesian(xlim=c(0, 1))
for (i in seq_along(geo.surrogacy.LST))
	p2 <- p2 + stat_smooth(method='glm', method.args=list(family='binomial'),
		data=filter(geo.surrogacy.DF, Species==unique(Species)[i]),
		fullrange=TRUE, se=FALSE, colour='black')

p1.hist <- ggplot(aes(r2CU),
		data=filter(env.surrogate.spp.DF, !is.na(r2CU))) +
	geom_histogram(fill='grey10', bins=5) +
	ylab('\nFrequency') + xlab(expression(Environmental~variables~(R^2))) +
	theme_classic() + theme(axis.line.x=element_line(color='black'),
		axis.line.y=element_line(color='black'),
		plot.background=element_rect(fill='transparent', colour='transparent')) +
		geom_text(x=0.05, y=6, label='(c)') +
		coord_cartesian(xlim=c(0, 1))

p2.hist <- ggplot(aes(r2CU),
		data=filter(geo.surrogate.spp.DF, !is.na(r2CU))) +
	geom_histogram(fill='grey10', bins=10) +
	ylab('\nFrequency') + xlab(expression(Geographic~variables~(R^2))) +
	theme_classic() + theme(axis.line.x=element_line(color='black'),
		axis.line.y=element_line(color='black'), 
		plot.background=element_rect(fill='transparent')) +
		geom_text(x=0.05, y=8, label='(d)') +
		coord_cartesian(xlim=c(0, 1))

# render figure
grid.arrange(p1,p2,p1.hist,p2.hist,nrow=2)
```

<!-- scenario analysis -->
To determine if the addition of environmental and geographic targets could actually improve the effectiveness of prioritisations, we generated prioritisations according to three scenarios using different targets (Figure 3). We then computed the proportion of adaptive and neutral genetic variation secured in each prioritisation (Figure 4). We found no evidence to suggest that any of the interaction terms were significant (${\chi^2}_{`r scenario.ANOVA.1.DF$Chi.Df[[2]]`} = `r round(scenario.ANOVA.1.DF$statistic[[2]],2)`$; $P `r pvalString(scenario.ANOVA.1.DF$Pr..Chisq.[[2]])`$). This result indicates the trends observed in each of the main effects were consistent across all other levels in the main effects.

First, we found that on average the prioritisations tended to secure more adaptive (mean `r round(mean(filter(scenario.sub.DF,Metric=='Adaptive variation')$genetic.held*100),2)` % secured in all prioritisations $\pm$ `r round(sd(filter(scenario.sub.DF,Metric=='Adaptive variation')$genetic.held*100),2)`) than neutral genetic variation (mean `r round(mean(filter(scenario.sub.DF,Metric=='Neutral variation')$genetic.held*100),2)` % secured in all prioritisations $\pm$ `r round(sd(filter(scenario.sub.DF,Metric=='Neutral variation')$genetic.held*100),2)`; ${\chi^2}_{`r scenario.ANOVA.3.DF$Chi.Df[[2]]`} = `r round(scenario.ANOVA.3.DF$statistic[[2]],2)`$; $P `r pvalString(scenario.ANOVA.3.DF$Pr..Chisq.[[2]])`$). This result indicates that regardless of the targets used to generate a prioritisation, or the planning context under which the prioritisation was generated, prioritisations tend to secure more adaptive then neutral genetic variation.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=(5*0.9), fig.height=(6.5*0.8), out.width='3.42in', height='3.9in', fig.cap='Planning unit selection frequencies. Squares represent planning units, and their colors indicate their selection frequency. Each column of panels shows solutions generated using different targets. Each row of panels shows solutions generated under different scenarios. Note that the selection frequencies for the single-species prioritizations are based on their frequency among prioritizations generated for each species. The selection frequencies for the multi-species prioritizations are each based on a single optimal pioritization for all species and so are binary.'}
# prepare data for plotting
grid.sub.FPLY <- grid.sub.PLY
for (i in seq_along(single.spp.prioritisations[[1]]))
	grid.sub.FPLY@data[[paste0('Single-species (equal costs).', c('Amount', 'Surrogate', 'Genetic')[i])]] <- rowMeans(sapply(single.spp.prioritisations, function(x){colMeans(x[[i]]@results@selections)}))
for (i in seq_along(single.spp.prioritisations.with.cost[[1]]))
	grid.sub.FPLY@data[[paste0('Single-species (opportunity costs).', c('Amount', 'Surrogate', 'Genetic')[i])]] <- rowMeans(sapply(single.spp.prioritisations.with.cost, function(x){colMeans(x[[i]]@results@selections)}))
for (i in seq_along(multi.spp.prioritisations))
	grid.sub.FPLY@data[[paste0('Multi-species (equal costs).', c('Amount', 'Surrogate', 'Genetic')[i])]] <- colMeans(multi.spp.prioritisations[[i]]@results@selections)
for (i in seq_along(multi.spp.prioritisations.with.cost))
	grid.sub.FPLY@data[[paste0('Multi-species (opportunity costs).', c('Amount', 'Surrogate', 'Genetic')[i])]] <- colMeans(multi.spp.prioritisations.with.cost[[i]]@results@selections)
grid.sub.FPLY <- spFortify(grid.sub.FPLY)
grid.sub.FPLY <- grid.sub.FPLY %>% gather(
		Combination,
		`Selection.Frequency`,
		`Single-species (equal costs).Amount`:`Multi-species (opportunity costs).Genetic`
	) %>% mutate(
		Context=sapply(strsplit(as.character(Combination), '.', fixed=TRUE), `[[`, 1),
		Prioritization=sapply(strsplit(as.character(Combination), '.', fixed=TRUE), `[[`, 2)
	) %>% mutate(
		Prioritization=revalue(Prioritization, c('Amount'='Amount targets', 'Surrogate'='Amount & surrogate\ntargets', 'Genetic'='Amount & genetic\ntargets')),
		Prioritization=factor(as.character(Prioritization), levels=c('Amount targets', 'Amount & surrogate\ntargets', 'Amount & genetic\ntargets')),
		Context=factor(
			gsub(' (', '\n(', Context, fixed=TRUE),
			levels=c("Single-species\n(equal costs)", "Single-species\n(opportunity costs)",
				"Multi-species\n(equal costs)", "Multi-species\n(opportunity costs)"))
	)

# make maps
ggplot() +
	geom_polygon(data=countries.FPLY, aes(x=long, y=lat, group=group),
		fill='grey85', color='grey70') +
	geom_polygon(data=grid.sub.FPLY, aes_string(x='long', y='lat', 
		group='group', fill='Selection.Frequency'),
		alpha=0.8, color=NA) +
	theme_classic() +
	theme(axis.ticks=element_blank(), axis.text=element_blank(),
		plot.margin=unit(c(0,0,0,0),'cm'), axis.line=element_blank(),
		strip.background = element_rect(fill='grey20'),
		strip.text = element_text(color='white', size=8),
		legend.position='bottom', legend.key.width=unit(1.8, 'cm'),
		legend.title=element_text(vjust=-5),
		legend.margin=unit(c(0.1), 'cm'),
		axis.title=element_blank()) +
	coord_cartesian(
		xlim=buffered.range(grid.sub.FPLY$long, 0),
		ylim=buffered.range(grid.sub.FPLY$lat, 0)
	) +
	facet_grid(Context ~ Prioritization) +
	scale_fill_gradientn(name='Selection Frequency (%)',
		colors=brewer.pal(9, 'PuBu'),
		guide=guide_colorbar(title.position='bottom', title.hjust=0.5,
			ticks=element_line(color='black'), border=element_line(color='black'))
	)
```

Second, the proportion of genetic variation secured in a prioritisation was found to depend on the combination of targets used to generate the prioritisation (${\chi^2}_{`r scenario.ANOVA.4.DF$Chi.Df[[2]]`} = `r round(scenario.ANOVA.4.DF$statistic[[2]],2)`$; $P `r pvalString(scenario.ANOVA.4.DF$Pr..Chisq.[[2]])`)$. Specifically, the amount-based prioritisations (`r round(mean(scenario.sub.DF,filter(Prioritisation=='Amount targets')$genetic.held*100),2)` % overall genetic variation secured $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount targets')$genetic.held*100),2)`) tended to secure less variation than the prioritisations generated using additional surrogate targets (`r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount & surrogate\ntargets')$genetic.held*100),2)` % secured $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount & surrogate\ntargets')$genetic.held*100),2)`; $Z_1 = `r round(scenario.GLHT.2.DF$statistic[[1]],2)`$; $P = `r pvalString(scenario.GLHT.2.DF$p.value[[1]])`$) or genetic targets (`r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount & genetic\ntargets')$genetic.held*100),2)` % secured $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount & genetic\ntargets')$genetic.held*100),2)`; `$Z_1 = `r round(scenario.GLHT.2.DF$statistic[[2]],2)`$; $P `r pvalString(scenario.GLHT.2.DF$p.value[[2]])`$). No evidence was found to suggest that the prioritisations that included additional surrogate-based targets (`r raptr.params.LST[[MODE]]$scenario.analysis$surrogate.target*100` %) or genetic-based targets (`r raptr.params.LST[[MODE]]$scenario.analysis$genetic.target*100` %) secured a different proportion of genetic variation ($Z_1 = `r round(scenario.GLHT.2.DF$statistic[[3]],2)`$; $P = `r pvalString(scenario.GLHT.2.DF$p.value[[3]])`$). These results suggest that the addition of environmental and geographic surrogates can improve the proportion of intra-specific genetic variation secured in a prioritisation for most species.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=(5*0.9), fig.height=(7.1*0.9), out.width='3.42in', out.height='4in', fig.cap='The effectiveness of prioritizations at securing genetic variation. Box plots show the proportion of adaptive and neutral genetic variation secured for each species using different targets and under different scenarios. Red lines show the target proportion of genetic variation to secure a representative sample.'}
# prepare data for plotting
scenario.SDF <- scenario.DF %>% 
	group_by(Prioritisation, Context, Metric, Species) %>% 
	summarize(genetic.held=mean(genetic.held, na.rm=TRUE)) %>%
	data.frame()
